---
title: "Chapter 1 HW"
author: "Orly Olbum"
date: 
output:
  pdf_document: default
  html_document: default
---

```{r setup, include = FALSE}

library(astsa)
library(xts)
```

## Homework 2 - Chapter 1

## Question 1.1

(a) Generate n=100 observations from autoregression xt = -.9xt-2 + wt with sigma(w) = 1, using Example 1.9. Next, apply moving average filter vt = (xt + xt-1 + xt-2 + xt-3)/4 to xt, the generated data. Now plot xt as a line and superimpose vt as a dashed line.

```{r}
par(mfrow = c(1, 1))
w = rnorm(150, 0 ,1) # 50 extra to avoid startup problems
xa = filter(w, filter = c(0, -.9), method = "recursive")[-(1:50)]
va = filter(xa, rep(1, 4)/4, sides = 1)
tsplot(xa, main = "autoregression part a")
lines(va, col = 2)
```

(b) Repeat (a) with xt = 2cos(2pit/4) + wt, where wt are iid ~N(0, 1).

```{r}
t = 1:100
w = rnorm(150, 0 ,1) # 50 extra to avoid startup problems
xb = 2*cos(2*pi*t/4)
vb = filter(xb, rep(1, 4)/4, sides = 1)
tsplot(xb, main = "autoregression part b")
lines(vb, col = 2)
```

(c) Repeat (a) where xt is log of Johnson & Johnson data from Example 1.1.

```{r}
w = rnorm(150, 0, 1) # 50 extra to avoid startup problems
xc = log(jj)
vc = filter(xc, rep(1, 4)/4, sides = 1)
tsplot(xc, main = "autoregression part c")
lines(vc, col = 2)
```

(d) What is seasonal adjustment?

Removal of the seasonal component of a time series - usually done when aiming to analyze a trend and cyclical deviations from a trend for time series separately from the seasonal components. Essentially each value of the time series is altered, whether by multiplication or by addition, by a seasonal index. Seasonal adjustment is the approximation of the seasonal component of a time series, and the removal of its effects to analyze a true trend.

(e) State conclusions.

We see in the examples in this question that the moving average takes away from the periodic aspect of the time series and allows us to analyze the mean.

## Question 1.3
In this problem, we explore difference between random walk and moving average models.

(a) Generate and plot nine series that are random walks (Example 1.10) of n = 500 and no drift, with sigma(w) = 1.

```{r}
par(mfrow = c(3, 3))
for(i in 1:9) {
  x = cumsum(rnorm(500))
  tsplot(x)
}
```

(b) Generate and plot nine series of n = 500 that are moving averages of form (1.1) in Example 1.8.

```{r}
par(mfrow = c(3, 3))
for(i in 1:9){
  y = filter(rnorm(500), filter = rep(1/9, 9))
  tsplot(y)
}
```

(c) Comment on differences between results from (a) and (b).

The random walks look more precise and tighter to the mean, while the moving averages are noisier.


## Question 1.4
Data in gdp are seasonally adjusted quarterly US GDP from 1947-I to 2018-III. Growth rate is shown in Figure 1.4.

(a) Plot data and compare it to a model discussed in Section 1.3.

```{r}
par(mfrow = c(1, 1))
tsplot(diff(log(gdp)), type = "o", col = 4, ylab = "GDP Growth")
points(diff(gdp)/lag(gdp, -1), pch = 3, col = 2)

# signal plus noise
t = 1:500
cs = log(gdp)
w = rnorm(500)
tsplot(cs, col = 4)

# autoregression
w = rnorm(150, 0, 1) # 50 extra to avoid startup problems
x = log(gdp)
v = filter(x, rep(1, 9)/9, sides = 1)
tsplot(x, main = "autoregression")
lines(v, col = 2)
```

(b) Reproduce Figure 1.4 in own colors and plot characters. Then, comment on difference between method of calculating growth rate.

```{r}
tsplot(diff(log(gdp)), type = "l", col = 3, ylab = "GDP Growth")
points(diff(gdp)/lag(gdp, -1), pch = 2, col = 6)

```

Logs: The growth rate calculated with logs equates to the percentage growth because change in logs are close to the percentage change in the actual data. Logging changes growth from multiplicative to additive (by the nature of logarithms). Difference between logs represents the percentage change from time period to time period.

Actual data: The growth rate calculated from actual data we are looking at is a multiplicative trend (or exponential).

(c) Which models discussed in Section 1.3 best describe behavior of US GDP growth?

The GPD data behaves like an autoregression - we are analyzing data that somewhat depends on the days before and after it (potentially - see "lag" in the modeling above). GPD potentially could fall into random walk as well.


