---
title: ""
author: ""
output:
  pdf_document: default
  word_document: default
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE)
library(astsa)
setwd("C:/Users/orlyo/OneDrive/Desktop/Grad School/Fall 2020/STAT 2321 - Applied Advanced Time Series")

```

## 2.11

(a) Simulate a series of n = 500 Gaussian white noise observations as in Example 1.7 and compute the sample ACF, p(h), to lag 20. Compare the sample ACF you obtain to the actual ACF, p(h). [Recall Example 2.17.]


```{r}
wa = rnorm(500)
va = filter(wa, side = 2, filter = rep(1/3, 3))
acf1(va, 20)

```

Compare sample ACF and true ACF:
Theoretically should be zero (actual ACF, calculated in the book), but some fluctuation around 0. Plotted, there are some residual values. Almost all spikes are within bounds so we can consider them white noise.


(b) Repeat part (a) using only n = 50. How does changing n affect the results?

```{r}
wb = rnorm(50)
vb = filter(wb, side = 2, filter = rep(1/3, 3))
acf1(vb, 20)

```

This changes results:
There is more variability because we have fewer observations, and we see a few more spikes beyond the bounds. We expect about 95% of observations to remain within the bounds.


## 2.12

(a) Simulate a series of n = 500 moving average observations as in Example 1.8 and compute the sample ACF, p(h), to lag 20. Compare the sample ACF you obtain to the actual ACF, p(h). [Recall Example 2.18 .]

```{r}

wa2 = rnorm(502, 0, 1)
va2 = filter(wa2, rep(1/3, 3))
acf1(va2, 20)

```

There are only two spikes beyond the bounds, at 1 and 2, otherwise they are within the bounds, whereas our actual ACF will be zero.

(b) Repeat part (a) using only n = 50. How does changing n affect the results?

```{r}
wb2 = rnorm(52, 0, 1)
vb2 = filter(wb2, rep(1/3, 3))
acf1(vb2, 20)

```

This changes the results: 
There is less variability with fewer observations generated, but we still have most observations within bounds leading our sample ACF to be approximately equal to the actual ACF.


## 2.13

Simulate 500 observations from the AR model specified in Example 1.9 and then plot the sample ACF to lag 50. What does the sample ACF tell you about the approximate cyclic behavior of the data? Hint: Recall Example 2.32.

xt = 1.5xt-1 - .75-x2 + wt

```{r}
set.seed(90210)
w = rnorm(500 + 50)
x = filter(w, filter = c(1.5, -.75), method = "recursive")[-(1:50)]
acf1(x, 50)

```

The sample ACF of the generated data with lag 50 shows cyclical behavior about every 10 units, with a positive autocorrelation exhibited on 5 units and negative autocorrelation every 5 units in a cyclical manner.
